---
title: "Untitled"
author: "Rossella Castronuovo"
date: "2025-10-14"
output: html_document
---

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

# Load libraries
library(readxl)
library(psych)
library(tidyverse)
library(FactoMineR)
library(magrittr)
library(ggplot2)
library(dplyr)
library(readr)
library(tools)
library(tidyr)

```

# Metrics selection and sector creation

All derived metrics were subsequently grouped into three structural domains:

1.	Tree sector: global size and stem architecture (e.g. DBH, height, total volume, trunk length).
2.	Branch sector: branching hierarchy and geometry (branch order, angles, lengths, volume fractions).
3.	Canopy sector: crown geometry and space occupation (crown length, crown volume, projected area, crown density).

This hierarchical approach separates strongly correlated variables, reduce dimensionality in a structured way, and avoid dominance of numerically redundant metric sets (Dormann et al. 2013). 

## 1. Analysis preparation
## 1.1 Input data

```{r message=FALSE, warning=TRUE, paged.print=FALSE}

# Load CSV cloud
QC_TR_Cloud<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Qc_Tr_cloud.csv")
QC_CP_Cloud<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Qc_Cp_cloud.csv")
PM_R_Cloud<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Pm_R_cloud.csv")
FS_HF_Cloud<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Fs_Hf_cloud.csv")

# Load branch file
QC_TR_Branch<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Qc_Tr_branch.csv")
QC_CP_Branch<- read_xlsx("E:/DELL/Lidar structure index/Stat/Results/Row_data/Qc_Cp_branch.xlsx")
PM_R_Branch<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Pm_R_branch.csv")
FS_HF_Branch<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Fs_Hf_branch.csv")
  
# Load mat file
QC_TR_TreeData<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Qc_Tr_treeData.csv")
QC_CP_TreeData <- read_delim("E:/DELL/Lidar structure index/Stat/Results/Row_data/Qc_Cp_treeData.csv",
                  delim = ";", 
                  show_col_types = FALSE)
PM_R_TreeData<- read_csv("E:/DELL/Lidar structure index/Stat/Results/Row_data/Pm_R_treeData.csv")
FS_HF_TreeData <- read_delim("E:/DELL/Lidar structure index/Stat/Results/Row_data/Fs_Hf_treeData.csv",
                  delim = ";",  
                  show_col_types = FALSE,  
                  trim_ws = TRUE)
  
```

## 1.2 Output data

```{r}
output <- "E:/DELL/Lidar structure index/Stat/Results/Sectors" 
```

## 1.3 Arrange TreeData sector

```{r}

# Extract treeData Sector Metrics from point-cloud

TreeSector <- function(cloud_df, 
                             pop_label = NULL) {
                                                  cloud_df %>%
                                                  dplyr::select(file_name, 
                                                                source, 
                                                                DAB, 
                                                                H   
) 
}

QC_TR_Tree <- TreeSector(QC_TR_Cloud, pop_label = "Qc_Tr")
QC_CP_Tree <- TreeSector(QC_CP_Cloud, pop_label = "Qc_Cp")
FS_HF_Tree <- TreeSector(FS_HF_Cloud, pop_label = "Fs_Hf")
PM_R_Tree <- TreeSector(PM_R_Cloud, pop_label = "PM_R")

# Extract treeData Sector Metrics from matfile

# --- PM_R ---
PM_R_TreeData <- PM_R_TreeData %>%
  select(-TreeHeight, -TotalArea) %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

PM_R_Cloud <- PM_R_Tree %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

PM_R_TreeDatasector <- PM_R_Cloud %>%
  left_join(select(PM_R_TreeData, -file_name), by = "key") %>%
  select(-key)

# --- Qc_Tr ---
QC_TR_TreeData <- QC_TR_TreeData %>%
  select(-TreeHeight, -TotalArea) %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

QC_TR_Cloud <- QC_TR_Tree %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

QC_TR_TreeDatasector <- QC_TR_Cloud %>%
  left_join(select(QC_TR_TreeData, -file_name), by = "key") %>%
  select(-key)

# --- Qc_Cp ---
QC_CP_TreeData <- QC_CP_TreeData %>%
  select(-TreeHeight, -TotalArea) %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

QC_CP_Cloud <- QC_CP_Tree %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

QC_CP_TreeDatasector <- QC_CP_Cloud %>%
  left_join(select(QC_CP_TreeData, -file_name), by = "key") %>%
  select(-key)

# --- Fs_Hf ---
FS_HF_TreeData <- FS_HF_TreeData %>%
  select(-TreeHeight, -TotalArea) %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

FS_HF_Cloud <- FS_HF_Tree %>%
  mutate(key = tolower(file_path_sans_ext(basename(file_name))))

FS_HF_TreeDatasector <- FS_HF_Cloud %>%
  left_join(select(FS_HF_TreeData, -file_name), by = "key") %>%
  select(-key)

#  Ba

QC_TR_TreeDatasector <- QC_TR_TreeDatasector %>%
  mutate(ba = pi * (DBHqsm/2)^2)

QC_CP_TreeDatasector <- QC_CP_TreeDatasector %>%
  mutate(ba = pi * (DBHqsm/2)^2)

FS_HF_TreeDatasector <- FS_HF_TreeDatasector %>%
  mutate(ba = pi * (DBHqsm/2)^2)

PM_R_TreeDatasector <- PM_R_TreeDatasector %>%
  mutate(ba = pi * (DBHqsm/2)^2)


# Save

write.csv(QC_TR_TreeDatasector, 
          file = file.path(output, "QC_TR_TreeSector.csv"), 
          row.names = FALSE)

write.csv(QC_CP_TreeDatasector, 
          file = file.path(output, "QC_CP_TreeSector.csv"), 
          row.names = FALSE)

write.csv(FS_HF_TreeDatasector, 
          file = file.path(output, "FS_HF_TreeSector.csv"), 
          row.names = FALSE)

write.csv(PM_R_TreeDatasector, 
          file = file.path(output, "PM_R_TreeSector.csv"), 
          row.names = FALSE)
```

## 1.4 Arrange Branch Data sector

```{r}

Branch_sector <- function(branch_df) {
  
  # ---- remove order 0 (trunk)
  branch_df <- branch_df %>%
    filter(order != 0) %>%
    rename(file_name = FileName)
  
  # ---- totals per tree
  totals <- branch_df %>%
    group_by(file_name) %>%
    summarise(
      totN = n(),
      totV = sum(volume, na.rm = TRUE),
      totL = sum(length, na.rm = TRUE),
      .groups = "drop"
    )
  
  # ---- metrics per tree × order
  long <- branch_df %>%
    group_by(file_name, order) %>%
    summarise(
      BOrd        = first(order),
      BN          = n(),
      
      # mean values
      BVol_mean   = mean(volume, na.rm = TRUE),
      BLen_mean   = mean(length, na.rm = TRUE),
      Bang_mean   = mean(angle,  na.rm = TRUE),
      
      # total values
      BVol_tot    = sum(volume,  na.rm = TRUE),
      BLen_tot    = sum(length,  na.rm = TRUE),
      
           .groups     = "drop"
    ) %>%
    left_join(totals, by = "file_name") %>%
    mutate(
      PN = if_else(totN > 0, 100 * BN        / totN, NA_real_),
      PV = if_else(totV > 0, 100 * BVol_tot  / totV, NA_real_),
      PL = if_else(totL > 0, 100 * BLen_tot  / totL, NA_real_)
    ) %>%
    arrange(file_name, BOrd)
  
  # ---- WIDE format
  wide <- long %>%
    pivot_wider(
      id_cols     = file_name,
      names_from  = BOrd,
      values_from = c(BN, BVol_mean, BVol_tot, 
                      BLen_mean, BLen_tot, Bang_mean, PN, PV, PL),
      names_glue  = "{.value}_o{BOrd}"
    )
  
  # ---- order columns (file_name first)
  ord_cols <- c("file_name", sort(setdiff(names(wide), "file_name")))
  wide[, ord_cols]
}

# APPLY 

QC_TR_BranchSector_wide <- Branch_sector(QC_TR_Branch)
QC_CP_BranchSector_wide <- Branch_sector(QC_CP_Branch)
FS_HF_BranchSector_wide <- Branch_sector(FS_HF_Branch)
PM_R_BranchSector_wide  <- Branch_sector(PM_R_Branch)

# SAVE

write.csv(QC_TR_BranchSector_wide, file.path(output, "QC_TR_BranchSector_WIDE.csv"), row.names = FALSE)
write.csv(QC_CP_BranchSector_wide, file.path(output, "QC_CP_BranchSector_WIDE.csv"), row.names = FALSE)
write.csv(FS_HF_BranchSector_wide, file.path(output, "FS_HF_BranchSector_WIDE.csv"), row.names = FALSE)
write.csv(PM_R_BranchSector_wide,  file.path(output, "PM_R_BranchSector_WIDE.csv"),  row.names = FALSE)

```

## 1.5 Arrange Canopy Data sector

```{r}

Canopy_sector <- function(tree_sector_df, cloud_df) {

  canopy_df <- cloud_df %>%
    transmute(
      file_name,
      source,
      crown_base_height,
      crown_length,
      crown_pa,
      crown_volume,
      H,
      cr_percent = 100 * (crown_length / H)
    )

  branchvol_df <- tree_sector_df %>%
    select(file_name, BranchVolume)   

  canopy_df %>%
    left_join(branchvol_df, by = "file_name") %>%
    mutate(
            crown_density = 100 * (BranchVolume / 1000) / crown_volume,    )
}

# APPLY TO ALL POPULATIONS

QC_TR_Canopy <- Canopy_sector(QC_TR_TreeDatasector, QC_TR_Cloud)
QC_CP_Canopy <- Canopy_sector(QC_CP_TreeDatasector, QC_CP_Cloud)
FS_HF_Canopy <- Canopy_sector(FS_HF_TreeDatasector, FS_HF_Cloud)
PM_R_Canopy <- Canopy_sector(PM_R_TreeDatasector, PM_R_Cloud)

# SAVE OUTPUT

QC_TR_Canopy %>%
  select(-BranchVolume, -H) %>%
  write.csv(file.path(output, "QC_TR_CanopySector.csv"), row.names = FALSE)

QC_CP_Canopy %>%
  select(-BranchVolume, -H) %>%
  write.csv(file.path(output, "QC_CP_CanopySector.csv"), row.names = FALSE)

FS_HF_Canopy %>%
  select(-BranchVolume, -H) %>%
  write.csv(file.path(output, "FS_HF_CanopySector.csv"), row.names = FALSE)

PM_R_Canopy %>%
  select(-BranchVolume, -H) %>%
  write.csv(file.path(output, "PM_R_CanopySector.csv"), row.names = FALSE)

```

