---
title: "Untitled"
author: "Rossella Castronuovo"
date: "2025-10-14"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

# Load Libraries
library(tidyverse)   
library(psych)
library(robustbase)
library(readxl)     
library(tools)
library(dplyr)
library(stringr)
library(rlang)

```

## 1. Analysis preparation

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

# Utility 
num_only  <- function(df) dplyr::select(df, where(is.numeric))
non_num   <- function(df) dplyr::select(df, !where(is.numeric))
get_clean <- function(x) if (is.list(x) && "clean" %in% names(x)) x$clean else x

```

## 1.1 Input data

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

# --- QC_TR ---
QC_TR_TreeSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/QC_TR_TreeSector.csv")%>%
  select(-DAB)
QC_TR_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/QC_TR_BranchSector_WIDE.csv")
QC_TR_CanopySector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/QC_TR_CanopySector.csv")

# --- QC_CP ---
QC_CP_TreeSector   <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/QC_CP_TreeSector.csv",)%>%
  select(-DAB)
QC_CP_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/QC_CP_BranchSector_WIDE.csv")
QC_CP_CanopySector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/QC_CP_CanopySector.csv")

# --- FS_HF ---
FS_HF_TreeSector   <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/FS_HF_TreeSector.csv")%>%
  select(-DAB)
FS_HF_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/FS_HF_BranchSector_WIDE.csv")
FS_HF_CanopySector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/FS_HF_CanopySector.csv")

# --- PM_R---
PM_R_TreeSector   <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/PM_R_TreeSector.csv")%>%
  select(-DAB)
PM_R_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/PM_R_BranchSector_WIDE.csv")
PM_R_CanopySector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/PM_R_CanopySector.csv")

```

## 1.2 Output data

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

output <- "C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors" 
dir.create(file.path(output, "01_descriptives"), recursive = TRUE, showWarnings = FALSE)

```

## 1.3 outlayer detection

Identifies and removes outliers from a data frame based on z-scores (the number of standard deviations from the mean) for all numeric variables.

1.  Selects only numeric columns from the input data frame.
2.  Computes the z-score for each numeric variable (mean = 0, sd = 1).
3.  Flags rows where at least one variable exceeds ±thresh (default = 3 SD).

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

z_outliers <- function(df, 
                       name, 
                       thresh = 3, 
                       output_dir = output) 
  
  {  cat("\nProcessing:", name, "\n")
  
     num_df <- dplyr::select(df, where(is.numeric))
     if (ncol(num_df) < 1) {
     warning("No numeric variables in ", name)
     return(df)
  }
  
z <- scale(num_df)
  
  # Flag  outlier (TRUE if > ±3 SD)
  df$is_outlier <- apply(abs(z) > thresh, 
                                   1,
                                   any, 
                                   na.rm = TRUE)
    n_out <- sum(df$is_outlier,
                 na.rm = TRUE)
  cat("Flagged", n_out, 
      "potential outliers (", 
      round(100 * n_out / nrow(df), 2), "%)\n")
  
  # save
  df_clean <- df[!df$is_outlier, ]
  readr::write_csv(df, file.path(output_dir, paste0(name, "_FLAGGED.csv")))
  readr::write_csv(df_clean, file.path(output_dir, paste0(name, "_CLEAN.csv")))
  
  invisible(list(clean = df_clean, outliers = df[df$is_outlier, ]))
}

# APPLY 

# QC_TR
QC_TR_Tree_clean   <- z_outliers(QC_TR_TreeSector,   "QC_TR_TreeSector")
QC_TR_Branch_clean <- z_outliers(QC_TR_BranchSector, "QC_TR_BranchSector")
QC_TR_Canopy_clean <- z_outliers(QC_TR_CanopySector, "QC_TR_CanopySector")

# QC_CP
QC_CP_Tree_clean   <- z_outliers(QC_CP_TreeSector,   "QC_CP_TreeSector")
QC_CP_Branch_clean <- z_outliers(QC_CP_BranchSector, "QC_CP_BranchSector")
QC_CP_Canopy_clean <- z_outliers(QC_CP_CanopySector, "QC_CP_CanopySector")

# FS_HF
FS_HF_Tree_clean   <- z_outliers(FS_HF_TreeSector,   "FS_HF_TreeSector")
FS_HF_Branch_clean <- z_outliers(FS_HF_BranchSector, "FS_HF_BranchSector")
FS_HF_Canopy_clean <- z_outliers(FS_HF_CanopySector, "FS_HF_CanopySector")

# PM_R
PM_R_Tree_clean    <- z_outliers(PM_R_TreeSector,    "PM_R_TreeSector")
PM_R_Branch_clean  <- z_outliers(PM_R_BranchSector,  "PM_R_BranchSector")
PM_R_Canopy_clean  <- z_outliers(PM_R_CanopySector,  "PM_R_CanopySector")

```

## Utilities: Volume unit conversion (L → m³)

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

scale_if_numeric <- function(df,cols,factor) 
  {
  present <- intersect(cols, names(df))
  if (length(present) == 0) return(df)

  df %>%
    mutate(across(all_of(present),
                  ~ if (is.numeric(.)) . / factor else .))
}

scale_if_numeric_regex <- function(df, pattern, factor) {
  target_cols <- names(df)[str_detect(names(df), pattern)]
  if (length(target_cols) == 0) return(df)
  
  df %>%
    mutate(across(all_of(target_cols),
                  ~ if (is.numeric(.)) . / factor else .))
}

# Tree sector: convert QSM volumes from L to m³
convert_tree_units <- function(df) {
  liter_cols <- c("TotalVolume", "TrunkVolume", "BranchVolume")
  scale_if_numeric(df, liter_cols, factor = 1000)
}

# Branch sector (wide): convert branch volumes from L to m³
convert_branch_units <- function(df) {
  pattern <- "^(BVol|BVT)"   # inizia con BVol o BVT
  scale_if_numeric_regex(df, pattern, factor = 1000)
}

# Canopy sector: crown_volume already in m³ (no change)
convert_canopy_units <- function(df) {
  df
}

# Wrapper: apply conversion to z_outliers() objects (list with $clean)
convert_clean_obj <- function(obj, converter) {
  if (is.list(obj) && "clean" %in% names(obj)) {
    obj$clean <- converter(obj$clean)
    return(obj)
  }
  converter(obj)
}


# Apply 

# TREE
QC_TR_Tree_clean <- convert_clean_obj(QC_TR_Tree_clean, convert_tree_units)
QC_CP_Tree_clean <- convert_clean_obj(QC_CP_Tree_clean, convert_tree_units)
FS_HF_Tree_clean <- convert_clean_obj(FS_HF_Tree_clean, convert_tree_units)
PM_R_Tree_clean  <- convert_clean_obj(PM_R_Tree_clean,  convert_tree_units)

# BRANCH
QC_TR_Branch_clean <- convert_clean_obj(QC_TR_Branch_clean, convert_branch_units)
QC_CP_Branch_clean <- convert_clean_obj(QC_CP_Branch_clean, convert_branch_units)
FS_HF_Branch_clean <- convert_clean_obj(FS_HF_Branch_clean, convert_branch_units)
PM_R_Branch_clean  <- convert_clean_obj(PM_R_Branch_clean,  convert_branch_units)

# CANOPY
QC_TR_Canopy_clean <- convert_clean_obj(QC_TR_Canopy_clean, convert_canopy_units)
QC_CP_Canopy_clean <- convert_clean_obj(QC_CP_Canopy_clean, convert_canopy_units)
FS_HF_Canopy_clean <- convert_clean_obj(FS_HF_Canopy_clean, convert_canopy_units)
PM_R_Canopy_clean  <- convert_clean_obj(PM_R_Canopy_clean,  convert_canopy_units)


save_clean_df <- function(obj, filename, output_dir = output) {
  
  # z_outliers() clean
  df_clean <- if (is.list(obj) && "clean" %in% names(obj)) obj$clean else obj
  
  # Save
  readr::write_csv(df_clean,
                   file.path(output_dir, paste0(filename, ".csv")))
  
}

# --- TREE ---
save_clean_df(QC_TR_Tree_clean,  "QC_TR_TreeSector_CLEAN")
save_clean_df(QC_CP_Tree_clean,  "QC_CP_TreeSector_CLEAN")
save_clean_df(FS_HF_Tree_clean,  "FS_HF_TreeSector_CLEAN")
save_clean_df(PM_R_Tree_clean,   "PM_R_TreeSector_CLEAN")

# --- BRANCH ---
save_clean_df(QC_TR_Branch_clean, "QC_TR_BranchSector_CLEAN")
save_clean_df(QC_CP_Branch_clean, "QC_CP_BranchSector_CLEAN")
save_clean_df(FS_HF_Branch_clean, "FS_HF_BranchSector_CLEAN")
save_clean_df(PM_R_Branch_clean,  "PM_R_BranchSector_CLEAN")

# --- CANOPY ---
save_clean_df(QC_TR_Canopy_clean, "QC_TR_CanopySector_CLEAN")
save_clean_df(QC_CP_Canopy_clean, "QC_CP_CanopySector_CLEAN")
save_clean_df(FS_HF_Canopy_clean, "FS_HF_CanopySector_CLEAN")
save_clean_df(PM_R_Canopy_clean,  "PM_R_CanopySector_CLEAN")

```

## 1.4 Descriptive Statistics

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

library(tidyverse)
library(psych)

# helpers
num_only  <- function(df) dplyr::select(df, where(is.numeric))
get_clean <- function(x) if (is.list(x) && "clean" %in% names(x)) x$clean else x

# describe() + CV(%)
describe_with_cv <- function(df) {
  df_num <- num_only(df)
  if (ncol(df_num) == 0) return(NULL)
  
  psych::describe(df_num) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Variable") %>%
    mutate(
      CV_perc = ifelse(
        is.finite(mean) & mean != 0,
        100 * sd / abs(mean),
        NA_real_
      )
    )
}

# TREE
QC_TR_Tree_df <- get_clean(QC_TR_Tree_clean)
QC_CP_Tree_df <- get_clean(QC_CP_Tree_clean)
FS_HF_Tree_df <- get_clean(FS_HF_Tree_clean)
PM_R_Tree_df  <- get_clean(PM_R_Tree_clean)

# BRANCH
QC_TR_Branch_df <- get_clean(QC_TR_Branch_clean)
QC_CP_Branch_df <- get_clean(QC_CP_Branch_clean)
FS_HF_Branch_df <- get_clean(FS_HF_Branch_clean)
PM_R_Branch_df  <- get_clean(PM_R_Branch_clean)

# CANOPY
QC_TR_Canopy_df <- get_clean(QC_TR_Canopy_clean)
QC_CP_Canopy_df <- get_clean(QC_CP_Canopy_clean)
FS_HF_Canopy_df <- get_clean(FS_HF_Canopy_clean)
PM_R_Canopy_df  <- get_clean(PM_R_Canopy_clean)

# Save 

output_desc <- file.path(output, "01_descriptives")
dir.create(output_desc, recursive = TRUE, showWarnings = FALSE)

save_desc <- function(df, pop, sector) {
  dsc <- describe_with_cv(df)
  if (is.null(dsc)) return(invisible(NULL))
  
  filename <- paste0("DESC_", pop, "_", sector, ".csv")
  readr::write_csv(dsc, file.path(output_desc, filename))
}

# TREE
save_desc(QC_TR_Tree_df, "QC_TR", "Tree")
save_desc(QC_CP_Tree_df, "QC_CP", "Tree")
save_desc(FS_HF_Tree_df, "FS_HF", "Tree")
save_desc(PM_R_Tree_df,  "PM_R",  "Tree")

# BRANCH
save_desc(QC_TR_Branch_df, "QC_TR", "Branch")
save_desc(QC_CP_Branch_df, "QC_CP", "Branch")
save_desc(FS_HF_Branch_df, "FS_HF", "Branch")
save_desc(PM_R_Branch_df,  "PM_R",  "Branch")

# CANOPY
save_desc(QC_TR_Canopy_df, "QC_TR", "Canopy")
save_desc(QC_CP_Canopy_df, "QC_CP", "Canopy")
save_desc(FS_HF_Canopy_df, "FS_HF", "Canopy")
save_desc(PM_R_Canopy_df,  "PM_R",  "Canopy")

```

```{r}
#LIST DATA

tree_list <- list(
  QC_TR = QC_TR_Tree_df,
  QC_CP = QC_CP_Tree_df,
  FS_HF = FS_HF_Tree_df,
  PM_R  = PM_R_Tree_df
)

branch_list <- list(
  QC_TR = QC_TR_Branch_df,
  QC_CP = QC_CP_Branch_df,
  FS_HF = FS_HF_Branch_df,
  PM_R  = PM_R_Branch_df
)

canopy_list <- list(
  QC_TR = QC_TR_Canopy_df,
  QC_CP = QC_CP_Canopy_df,
  FS_HF = FS_HF_Canopy_df,
  PM_R  = PM_R_Canopy_df
)

# output
output_desc <- file.path(output, "01_descriptives")

make_sector_summary <- function(sector_list, sector_name) {
  all_sec <- purrr::imap_dfr(
    sector_list,
    function(df, pop) {
      dsc <- describe_with_cv(df)
      if (is.null(dsc)) return(NULL)
      dsc %>%
        mutate(
          Population = pop,
          Sector     = sector_name,
          .before = 1  
        )
    }
  )
  
  # SAVE
  filename <- paste0("DESC_ALL_", sector_name, ".csv")
  readr::write_csv(all_sec, file.path(output_desc, filename))
  message("Saved: ", filename)
  
  all_sec
}

# APPLY

DESC_ALL_Tree   <- make_sector_summary(tree_list,   "Tree")
DESC_ALL_Branch <- make_sector_summary(branch_list, "Branch")
DESC_ALL_Canopy <- make_sector_summary(canopy_list, "Canopy")

```

## 1.4.1 Descriptive Plots for tree sector

```{r}
# ============================================================
# Descriptive barplots — TREE sector
# Requested edits:
# 1) Title: "Average <full variable name> per tree"
# 2) Axis titles NOT bold
# 3) Keep y-axis label with ORIGINAL variable name (and units)
# 4) Save each single plot + ONE final multi-panel image
# 5) Final multi-panel at 1200 dpi
# ============================================================

library(dplyr)
library(ggplot2)
library(patchwork)

# ---------------------------
# UNITS (keep your originals)
# ---------------------------
tree_var_units <- c(
  H            = "m",
  ba           = "m²",
  TotalVolume  = "m³",
  TrunkVolume  = "m³",
  BranchVolume = "m³",
  BranchLength = "m",
  TrunkLength  = "m",
  DBHqsm       = "m"
)

# -------------------------------------------------
# FULL (extended) names for titles (edit as needed)
# -------------------------------------------------
tree_var_full <- c(
  H            = "Tree height",
  ba           = "Basal area",
  TotalVolume  = "Total volume",
  TrunkVolume  = "Trunk volume",
  BranchVolume = "Branch volume",
  BranchLength = "Total branch length",
  MaxBranchOrder = "Max Branch Order",
  TrunkLength  = "Trunk length",
  DBHqsm       = "Diameter at breast height",
  NumberBranches = "Number of Branches"
)

# folder plots output
tree_plot_dir <- file.path(output, "02_plots_Tree_bar")
dir.create(tree_plot_dir, recursive = TRUE, showWarnings = FALSE)

make_ylab <- function(v)
  if (v %in% names(tree_var_units))
    paste0(v, " [", tree_var_units[[v]], "]") else v

make_title <- function(v)
  paste0("Average ",
         if (v %in% names(tree_var_full)) tree_var_full[[v]] else v,
         " per tree")

# ---------------------------
# SINGLE PLOT
# ---------------------------
plot_tree_variable <- function(var_name, tag_letter,
                               desc_tbl = DESC_ALL_Tree,
                               dpi_single = 600) {

  dat <- desc_tbl %>% filter(Variable == var_name)
  if (nrow(dat) == 0) return(NULL)

  p <- ggplot(dat, aes(x = Population, y = mean, fill = Population)) +
    geom_col(color = "black", width = 0.65) +
    geom_errorbar(
      aes(ymin = mean - sd, ymax = mean + sd),
      width = 0.2, color = "black"
    ) +
    labs(
      title = make_title(var_name),
      x = "Forest stands",
      y = make_ylab(var_name),
      tag = tag_letter
    ) +
    scale_fill_grey(start = 0.85, end = 0.2) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5),     # COME PRIMA
      axis.title.x = element_text(face = "plain"),
      axis.title.y = element_text(face = "plain"),

      # bordo sottile
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.4),

      # lettera in basso a destra
      plot.tag.position = c(0.98, 0.02),
      plot.tag = element_text(
        face = "plain",
        size = 12,
        hjust = 1,
        vjust = 0
      )
    )

  ggsave(
    filename = file.path(tree_plot_dir, paste0("BAR_Tree_", var_name, ".png")),
    plot = p, width = 7, height = 5, dpi = dpi_single
  )

  p
}

# ---------------------------
# BUILD ALL PLOTS WITH TAGS
# ---------------------------
tree_vars <- unique(DESC_ALL_Tree$Variable)
letters_vec <- letters[seq_along(tree_vars)]

plot_list <- Map(
  plot_tree_variable,
  var_name   = tree_vars,
  tag_letter = letters_vec
)

plot_list <- Filter(Negate(is.null), plot_list)

# ---------------------------
# FINAL MULTI-PANEL
# ---------------------------
ncol_final <- 2

final_plot <- wrap_plots(plot_list, ncol = ncol_final)

ggsave(
  filename = file.path(tree_plot_dir, "BAR_Tree_ALL_metrics_grid.png"),
  plot   = final_plot,
  width  = 7 * ncol_final,
  height = 5 * ceiling(length(plot_list) / ncol_final),
  dpi    = 1200
)

final_plot
```


## 1.4.2 Descriptive Plots for branch sector

```{r}

library(stringr)

#  folder plots output 
branch_plot_dir <- file.path(output, "02_plots_Branch_bar")
dir.create(branch_plot_dir, recursive = TRUE, showWarnings = FALSE)

# arrange data
branch_desc_long <- DESC_ALL_Branch %>%
  mutate(
    Metric = str_remove(Variable, "_o[0-9]+$"),                 # "BN_o1" -> "BN"
    Order  = as.integer(str_extract(Variable, "(?<=_o)[0-9]+")) # "BN_o1" -> 1
  ) %>%
  filter(!is.na(Order))

# label
branch_y_lab <- function(metric) {
  dplyr::case_when(
    metric == "BN"          ~ "Mean number of branches per tree",
    metric == "BLen_mean"   ~ "Mean branch length [m]",
    metric == "BLen_tot"    ~ "Total branch length [m]",
    metric == "BVol_mean"   ~ "Mean branch volume [m³]",
    metric == "BVol_tot"    ~ "Total branch volume [m³]",
    metric == "Bang_mean"   ~ "Mean branch angle [deg]",
    TRUE                    ~ metric
  )
}

# plot
plot_branch_metric <- function(metric, desc_tbl = branch_desc_long) {
  
  dat <- desc_tbl %>% filter(Metric == metric)
  if (nrow(dat) == 0) return(invisible(NULL))
 
  ord_levels <- sort(unique(dat$Order))
  dat <- dat %>% mutate(Order = factor(Order, levels = ord_levels))
  
  p <- ggplot(dat, aes(x = Order,
                       y = mean,
                       fill = Population)) +
    geom_col(position = position_dodge(width = 0.7),
             color = "black",
             width = 0.7) +
    geom_errorbar(
      aes(ymin = mean - sd, ymax = mean + sd),
      position = position_dodge(width = 0.7),
      width = 0.2,
      color = "black"
    ) +
    labs(
      title = paste0(metric, " – Branch sector"),
      x = "Branch order",
      y = branch_y_lab(metric),
      fill = "Population"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      panel.grid = element_blank(),
      panel.background = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title.x = element_text(face = "bold"),
      axis.title.y = element_text(face = "bold")
    ) +
    scale_fill_grey(start = 0.85, end = 0.2)
  
  ggsave(
    filename = file.path(branch_plot_dir,
                         paste0("BAR_Branch_", metric, "_byOrder.png")),
    plot   = p,
    width  = 7,
    height = 5,
    dpi    = 600
  )
  
  invisible(p)
}

branch_metrics <- unique(branch_desc_long$Metric)

# save
invisible(lapply(branch_metrics, plot_branch_metric))

# summary table

branch_summary_table <- branch_desc_long %>%
  select(Population, Metric, Order, n, mean, sd, CV_perc) %>%
  arrange(Metric, Order, Population)

# save
output_summary <- file.path(output, "01_descriptives")
readr::write_csv(branch_summary_table,
                 file.path(output_summary, "BranchSector_SUMMARY.csv"))

# stampa anteprima
print(head(branch_summary_table, 20))


```

## 1.4.2 Descriptive Plots for canopy sector

```{r}

# ============================================================
# Descriptive barplots — CANOPY sector
# Fix: crown_density plotted on 0–100 scale (like other % vars)
# ============================================================

library(dplyr)
library(ggplot2)
library(patchwork)

# UNITS
canopy_var_units <- c(
  crown_base_height = "m",
  crown_length      = "m",
  crown_pa          = "m²",
  crown_volume      = "m³",
  cr_percent        = "%",
  crown_density     = "%"
)

# FULL names
canopy_var_full <- c(
  crown_base_height = "Crown base height",
  crown_length      = "Crown length",
  crown_pa          = "Crown projected area",
  crown_volume      = "Crown volume",
  cr_percent        = "Crown ratio",
  crown_density     = "Crown density"
)

canopy_plot_dir <- file.path(output, "02_plots_Canopy_bar")
dir.create(canopy_plot_dir, recursive = TRUE, showWarnings = FALSE)

make_ylab_canopy <- function(v) {
  if (v %in% names(canopy_var_units)) paste0(v, " [", canopy_var_units[[v]], "]") else v
}

make_title_canopy <- function(v) {
  full <- if (v %in% names(canopy_var_full)) canopy_var_full[[v]] else v
  paste0("Average ", full, " per tree")
}

plot_canopy_variable <- function(var_name, tag_letter,
                                 desc_tbl = DESC_ALL_Canopy,
                                 dpi_single = 600) {

  dat <- desc_tbl %>% filter(Variable == var_name)
  if (nrow(dat) == 0) return(NULL)

  # --- FIX: crown_density -> percent for plotting (mean and sd) ---
  if (var_name == "crown_density") {
    dat <- dat %>%
      mutate(
        mean = mean * 100,
        sd   = sd   * 100
      )
  }

  p <- ggplot(dat, aes(x = Population, y = mean, fill = Population)) +
    geom_col(color = "black", width = 0.65) +
    geom_errorbar(
      aes(ymin = mean - sd, ymax = mean + sd),
      width = 0.2,
      color = "black"
    ) +
    labs(
      title = make_title_canopy(var_name),
      x = "Forest stands",
      y = make_ylab_canopy(var_name),
      tag = tag_letter
    ) +

    # --- FIX: force 0–100 scale for crown_density (like % variables) ---
    { if (var_name == "crown_density") scale_y_continuous(limits = c(0, 100)) } +

    theme_minimal(base_size = 13) +
    theme(
      panel.grid = element_blank(),
      legend.position = "none",
      plot.title   = element_text(hjust = 0.5),
      axis.title.x = element_text(face = "plain"),
      axis.title.y = element_text(face = "plain"),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.4),
      plot.tag.position = c(0.98, 0.02),
      plot.tag = element_text(face = "plain", size = 12, hjust = 1, vjust = 0)
    ) +
    scale_fill_grey(start = 0.85, end = 0.2)

  ggsave(
    filename = file.path(canopy_plot_dir, paste0("BAR_Canopy_", var_name, ".png")),
    plot   = p,
    width  = 7,
    height = 5,
    dpi    = dpi_single
  )

  p
}

# Build + save all single plots
canopy_vars <- unique(DESC_ALL_Canopy$Variable)
letters_vec <- letters[seq_along(canopy_vars)]

plot_list <- Map(plot_canopy_variable, var_name = canopy_vars, tag_letter = letters_vec)
plot_list <- Filter(Negate(is.null), plot_list)

# Final multi-panel
ncol_final <- 2
final_plot <- wrap_plots(plot_list, ncol = ncol_final)

ggsave(
  filename = file.path(canopy_plot_dir, "BAR_Canopy_ALL_metrics_grid.png"),
  plot   = final_plot,
  width  = 7 * ncol_final,
  height = 5 * ceiling(length(plot_list) / ncol_final),
  dpi    = 1200
)

final_plot


```
