---
title: "Untitled"
author: "Rossella Castronuovo"
date: "2025-10-14"
output: pdf_document
editor_options: 
markdown: 
wrap: 72
---

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

library(dplyr)
library(tidyr)
library(corrplot)
library(readr)
library(tibble)
library(ggplot2)
library(psych)
library(corrplot)
library(readr)

```

# Sector-level PCA analysis
A hierarchical multivariate approach was used to describe structural heterogeneity while preserving the ecological meaning of each structural sector. 

# Input files
```{r message=FALSE, warning=FALSE, paged.print=FALSE}

# Load Sector Files (Tree, Branch, Canopy) 

# QC_TR 
QC_TR_TreeSector   <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/QC_TR_TreeSector_CLEAN.csv")
QC_TR_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/QC_TR_BranchSector_CLEAN.csv")
QC_TR_CanopySector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/QC_TR_CanopySector_CLEAN.csv")

# QC_CP
QC_CP_TreeSector   <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/QC_CP_TreeSector_CLEAN.csv")
QC_CP_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/QC_CP_BranchSector_CLEAN.csv")
QC_CP_CanopySector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/QC_CP_CanopySector_CLEAN.csv")

# FS_HF 
FS_HF_TreeSector   <- read_delim ("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/FS_HF_TreeSector_CLEAN.csv", 
  delim = ",",  show_col_types = FALSE,  trim_ws = TRUE)
FS_HF_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/FS_HF_BranchSector_CLEAN.csv")
FS_HF_CanopySector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/FS_HF_CanopySector_CLEAN.csv")

# PM_R
PM_R_TreeSector   <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/PM_R_TreeSector_CLEAN.csv")
PM_R_BranchSector <- read_csv("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/PM_R_BranchSector_CLEAN.csv")
PM_R_CanopySector <- read_csv ("C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/Sectors/Clean_sectors/PM_R_CanopySector_CLEAN.csv")

```

## Output data

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
output <- "C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/PCA_Sectors" 
output2 <- "C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/PCA_Sectors/02_pcasector_matrices" 
output3 <- "C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/PCA_Sectors/011_drop_flag_var"

```

#  ADD Population Label

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

#  ID
id_col <- "file_name"

# List sectors per population
Tree_list <- list(
  QC_TR = QC_TR_TreeSector,
  QC_CP = QC_CP_TreeSector,
  FS_HF = FS_HF_TreeSector,
  PM_R  = PM_R_TreeSector
)

Branch_list <- list(
  QC_TR = QC_TR_BranchSector,
  QC_CP = QC_CP_BranchSector,
  FS_HF = FS_HF_BranchSector,
  PM_R  = PM_R_BranchSector
)

Canopy_list <- list(
  QC_TR = QC_TR_CanopySector,
  QC_CP = QC_CP_CanopySector,
  FS_HF = FS_HF_CanopySector,
  PM_R  = PM_R_CanopySector
)

```

# Correlation Matrix

For each sector and forest stand, Pearson correlation matrices were computed using “corrplot” package (Wei et al. 2024) in R. Variables involved in highly correlated pairs were progressively removed using sector-specific correlation thresholds (|r| ≥ 0.70 for Tree and Canopy; |r| ≥ 0.60 for Branch), chosen to account for differences in metric redundancy among structural domains.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

cor_output_dir <- file.path(output, "01_Correlation_matrices")
dir.create(cor_output_dir, recursive = TRUE, showWarnings = FALSE)

# List files
Tree_clean_list   <- list()
Branch_clean_list <- list()
Canopy_clean_list <- list()
Tree_cor_list     <- list()
Canopy_cor_list   <- list()
drop_log_all      <- list()

# Core function
reduce_cor_vars <- function(df,
                            sector,
                            pop,
                            threshold = 0.7,
                            out_dir   = cor_output_dir) {

  # numeric only
  df_num <- df %>% select(where(is.numeric))

  # drop zero-variance / all-NA
  vars_sd <- apply(df_num, 2, sd, na.rm = TRUE)
  ok_cols <- is.finite(vars_sd) & vars_sd > 0
  df_num  <- df_num[, ok_cols, drop = FALSE]

  # correlation
  cor_mat <- cor(df_num, use = "pairwise.complete.obs")
  cor_mat[is.na(cor_mat)] <- 0

  # save matrix csv
  cor_df <- as.data.frame(cor_mat) %>% rownames_to_column("Variable")
  write_csv(cor_df, file.path(out_dir, paste0("COR_", sector, "_", pop, ".csv")))

  # save single matrix png 
  png(file.path(out_dir, paste0("COR_", sector, "_", pop, ".png")),
      width = 1600, height = 1400, res = 150)

  corrplot(cor_mat,
           method      = "color",
           type        = "lower",
           addCoef.col = "black",
           number.cex  = 0.6,
           tl.cex      = 0.7,
           tl.col      = "black",
           mar         = c(0, 0, 0, 0))
  mtext(paste0(sector, " \u2013 ", pop), side = 3, line = 0.8, font = 1, cex = 1.1)
  dev.off()

  # iterative dropping
  to_drop   <- character()
  drop_info <- tibble()
  cm        <- cor_mat

  repeat {
    cor_abs <- abs(cm)
    diag(cor_abs) <- 0
    max_cor <- max(cor_abs)

    if (!is.finite(max_cor) || max_cor < threshold || ncol(cm) < 2) break

    pair <- which(cor_abs == max_cor, arr.ind = TRUE)[1, ]
    i <- pair[1]; j <- pair[2]

    mean_i <- mean(cor_abs[i, ], na.rm = TRUE)
    mean_j <- mean(cor_abs[j, ], na.rm = TRUE)

    drop_idx <- if (mean_i >= mean_j) i else j
    keep_idx <- if (drop_idx == i) j else i

    var_drop <- colnames(cm)[drop_idx]
    var_keep <- colnames(cm)[keep_idx]

    drop_info <- bind_rows(
      drop_info,
      tibble(
        Sector           = sector,
        Population       = pop,
        Dropped_var      = var_drop,
        Paired_with      = var_keep,
        Max_corr_at_drop = round(max_cor, 3),
        Signed_corr      = round(cor_mat[var_drop, var_keep], 3),
        Threshold_used   = threshold
      )
    )

    to_drop <- c(to_drop, var_drop)
    cm <- cm[-drop_idx, -drop_idx, drop = FALSE]
  }

  keep <- setdiff(colnames(df_num), to_drop)

  list(
    df_reduced = df_num[, keep, drop = FALSE],
    dropped    = drop_info,
    cor_full   = cor_mat
  )
}

# Run: TREE

for (pop in names(Tree_list)) {

  df     <- Tree_list[[pop]]
  id_vec <- df[[id_col]]

  res <- reduce_cor_vars(
    df        = df,
    sector    = "Tree",
    pop       = pop,
    threshold = 0.7
  )

  Tree_cor_list[[pop]] <- res$cor_full

  Tree_clean_list[[pop]] <- cbind(
    file_name = id_vec,
    res$df_reduced
  )

  if (nrow(res$dropped) > 0) drop_log_all[[paste0("Tree_", pop)]] <- res$dropped
}


# Run: BRANCH 

for (pop in names(Branch_list)) {

  df     <- Branch_list[[pop]]
  id_vec <- df[[id_col]]

  res <- reduce_cor_vars(
    df        = df,
    sector    = "Branch",
    pop       = pop,
    threshold = 0.6
  )

  Branch_clean_list[[pop]] <- cbind(
    file_name = id_vec,
    res$df_reduced
  )

  if (nrow(res$dropped) > 0) drop_log_all[[paste0("Branch_", pop)]] <- res$dropped
}

# Run: CANOPY

for (pop in names(Canopy_list)) {

  df     <- Canopy_list[[pop]]
  id_vec <- df[[id_col]]

  res <- reduce_cor_vars(
    df        = df,
    sector    = "Canopy",
    pop       = pop,
    threshold = 0.7
  )

  Canopy_cor_list[[pop]] <- res$cor_full

  Canopy_clean_list[[pop]] <- cbind(
    file_name = id_vec,
    res$df_reduced
  )

  if (nrow(res$dropped) > 0) drop_log_all[[paste0("Canopy_", pop)]] <- res$dropped
}


# Save drop logs (all sectors)

if (length(drop_log_all) > 0) {

  drop_log_df <- bind_rows(drop_log_all, .id = "Sector_Pop") %>%
    arrange(Sector, Population, desc(Max_corr_at_drop))

  write_csv(drop_log_df, file.path(cor_output_dir, "Drop_log_multicollinearita_all_sectors.csv"))
  write_csv(drop_log_df, file.path(cor_output_dir, "Dropped_variables_summary.csv"))
}

# 2x2 GLOBAL PANELS (TREE + CANOPY)

save_corrpanel_2x2 <- function(cor_list,
                               pops_order  = c("QC_TR","QC_CP","FS_HF","PM_R"),
                               sector_label = "Tree",
                               out_file,
                               res_dpi = 1200) {

  w_in <- 12
  h_in <- 10

  png(out_file, width = w_in * res_dpi, height = h_in * res_dpi, res = res_dpi)

  op <- par(no.readonly = TRUE)
  on.exit({par(op); dev.off()}, add = TRUE)

   par(mfrow = c(2, 2),
      mar = c(2.0, 4.2, 2.6, 2.0),  
      oma = c(0, 0, 0, 0))

  tags <- letters[seq_along(pops_order)]

  for (k in seq_along(pops_order)) {

    pop <- pops_order[k]
    cm  <- cor_list[[pop]]
    if (is.null(cm) || length(cm) == 0) next

    corrplot(cm,
             method      = "color",
             type        = "lower",
             addCoef.col = "black",
             number.cex  = 0.55,
             tl.cex      = 0.75,
             tl.col      = "black",
             mar         = c(0, 0, 0, 0))

  
    mtext(paste0(sector_label, " \u2013 ", pop), side = 3, line = 0.9, font = 1, cex = 1.0)
    usr <- par("usr")
    text(x = usr[2], y = usr[3], labels = tags[k],
         adj = c(1, 0), cex = 1.0, font = 1, xpd = NA)
  }
}

cor_panel_dir <- cor_output_dir

save_corrpanel_2x2(
  cor_list      = Tree_cor_list,
  pops_order    = c("QC_TR","QC_CP","FS_HF","PM_R"),
  sector_label  = "Tree",
  out_file      = file.path(cor_panel_dir, "COR_Tree_ALL_2x2.png"),
  res_dpi       = 1200
)

save_corrpanel_2x2(
  cor_list      = Canopy_cor_list,
  pops_order    = c("QC_TR","QC_CP","FS_HF","PM_R"),
  sector_label  = "Canopy",
  out_file      = file.path(cor_panel_dir, "COR_Canopy_ALL_2x2.png"),
  res_dpi       = 1200
)


```


# Correlation Matrix cvs

```{r}
#   LIST OF KEPT / DROPPED VARIABLES

#  Sector lists 
sector_lists <- list(
  Tree   = Tree_clean_list,
  Branch = Branch_clean_list,
  Canopy = Canopy_clean_list
)

# Final containers 
kept_vars_all    <- list()
dropped_vars_all <- list()


# Extract kept and dropped variables by sector and population ---
for (sector in names(sector_lists)) {

  cat("\n=== SECTOR:", sector, "===\n")
  pop_list <- sector_lists[[sector]]

  for (pop in names(pop_list)) {

    cat(" - Population:", pop, "\n")
    df_clean <- pop_list[[pop]]

    # Numerical variables in the cleaned dataset
    vars_all <- names(df_clean)[sapply(df_clean, is.numeric)]

    # Variables dropped due to multicollinearity (from log)
    vars_dropped <- drop_log_df %>%
      filter(Sector == sector, Population == pop) %>%
      pull(Dropped_var) %>%
      unique()

    # Variables retained
    vars_kept <- setdiff(vars_all, vars_dropped)

    # Store results
    key <- paste0(sector, "_", pop)
    kept_vars_all[[key]]    <- vars_kept
    dropped_vars_all[[key]] <- vars_dropped
  }
}

# Data frames ready for CSV export
kept_df <- tibble(
  Sector_Population = names(kept_vars_all),
  Kept_variables    = vapply(
    kept_vars_all,
    function(x) paste(x, collapse = "; "),
    character(1)
  )
)

dropped_df <- tibble(
  Sector_Population = names(dropped_vars_all),
  Dropped_variables = vapply(
    dropped_vars_all,
    function(x) paste(x, collapse = "; "),
    character(1)
  )
)

# --- 5) EXPORT (fix: create output directory before writing files) ---
dir.create(output3, recursive = TRUE, showWarnings = FALSE)

file_kept    <- file.path(output3, "Kept_variables_by_sector_population.csv")
file_dropped <- file.path(output3, "Dropped_variables_by_sector_population.csv")

write_csv(kept_df, file_kept)
write_csv(dropped_df, file_dropped)

cat(
  "\nFiles successfully exported:\n",
  "- ", file_kept, "\n",
  "- ", file_dropped, "\n"
)

```
##  PCA PER SECTOR

The resulting variable sets were used in an independent Principal Component Analyses (PCA) performed for each sector in a given forest stand using R (“stats” package). Principal components were retained until at least 60% of cumulative variance was explained; when PC1 alone exceeded this threshold, PC2 was additionally retained to avoid oversimplification of within-sector structural variability. For each forest stand, a sector-level indicators were computed as a variance-weighted linear combination of the retained component scores 

```{r}

run_sector_pca <- function(clean_list,
                           sector,               
                           id_col = "file_name",
                           min_var_PC1 = 0.60,
                           results_dir = "output2",
                           plot_dir    = "output2") {
  
  dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(plot_dir,    recursive = TRUE, showWarnings = FALSE)
  
  scores_all <- list()
  var_all    <- list()
  load_all   <- list()
  
  for (pop in names(clean_list)) {
    
    df <- clean_list[[pop]]
    
    # 1. Numerical vars
    num <- df %>% select(where(is.numeric))
    
    # 2. Replace Inf/-Inf → NA
    num[!is.finite(as.matrix(num))] <- NA
    
    # 3. Remove variables with zero variance
    sd_vec <- apply(num, 2, sd, na.rm = TRUE)
    num <- num[, sd_vec > 0, drop = FALSE]
    
    if (ncol(num) < 2) {
      warning("Popolamento ", pop, ": less then two var.")
      next
    }
    
    # 4. Remove rows with NA/NaN/Inf → only complete cases
    good_rows <- complete.cases(num)
    
    if (sum(good_rows) < 2) {
      warning("Popolamento ", pop, ": less then two var.")
      next
    }
    
    num_clean <- num[good_rows, , drop = FALSE]
    id_vec    <- df[[id_col]][good_rows]
    
    # 5. PCA
    pca <- prcomp(num_clean, center = TRUE, scale. = TRUE)
    
    var_expl <- (pca$sdev^2) / sum(pca$sdev^2)
    cum_var  <- cumsum(var_expl)
    
    keep_PC2 <- var_expl[1] < min_var_PC1 && length(var_expl) >= 2
    
    # 6. Scores
    score_df <- as.data.frame(pca$x)
    colnames(score_df) <- paste0(sector, "_PC", seq_len(ncol(score_df)))
    
    keep_cols <- paste0(sector, "_PC1")
    if (keep_PC2) keep_cols <- c(keep_cols, paste0(sector, "_PC2"))
    
    score_df <- score_df[, keep_cols, drop = FALSE]
    score_df[[id_col]] <- id_vec
    score_df$Pop <- pop
    scores_all[[pop]] <- score_df
    
    # 7. Variance explained
    var_df <- tibble(
      Pop        = pop,
      Component  = paste0("PC", seq_along(var_expl)),
      VarExpl    = var_expl,
      CumVarExpl = cum_var
    )
    var_all[[pop]] <- var_df
    
    # 8. Loadings
    load_df <- as.data.frame(pca$rotation)
    load_df$Variable <- rownames(load_df)
    rownames(load_df) <- NULL
    load_df <- load_df %>% relocate(Variable)
    load_df$Pop <- pop
    load_all[[pop]] <- load_df
    
    # 9. Scree plot

var_df$Component <- factor(
  var_df$Component,
  levels = paste0("PC", seq_len(nrow(var_df)))  # ordina PC1, PC2, PC3, ...
)

scree <- ggplot(var_df, aes(x = Component, y = VarExpl)) +
  geom_col(fill = "grey40") +
  geom_line(aes(group = 1)) +
  geom_point(size = 2) +
  theme_minimal(base_size = 14) + 
  theme(
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = paste0(sector, " - ", pop, " - Scree Plot"),
    x = "Component",
    y = "Variance Explained"
  )

ggsave(
  filename = file.path(plot_dir, paste0("Scree_", sector, "_", pop, ".png")),
  plot = scree,
  width = 7, height = 5, dpi = 300
)

    # 10. Score plot PC1 vs PC2
    if (ncol(pca$x) >= 2) {
      sp <- as.data.frame(pca$x)
      colnames(sp)[1:2] <- c("PC1", "PC2")
      sp[[id_col]] <- id_vec
      
      g <- ggplot(sp, aes(PC1, PC2)) +
        geom_point(alpha = 0.7) +
        labs(title = paste0(sector, " - ", pop, " - Scores"),
             x = "PC1", y = "PC2")
      
      ggsave(file.path(plot_dir, paste0("Scores_", sector, "_", pop, ".png")),
             g, width = 6, height = 4, dpi = 300)
    }
  }
  
  list(
    scores   = bind_rows(scores_all),
    variance = bind_rows(var_all),
    loadings = bind_rows(load_all)
  )
}

## PCA per Sector

res_tree <- run_sector_pca(
  clean_list  = Tree_clean_list,
  sector      = "Tree",
  id_col      = "file_name",
  min_var_PC1 = 0.60,
  results_dir = "output2",
  plot_dir    = "output2"
)

res_branch <- run_sector_pca(
  clean_list  = Branch_clean_list,
  sector      = "Branch",
  id_col      = "file_name",
  min_var_PC1 = 0.60,
  results_dir = "output2",
  plot_dir    = "output2"
)

res_canopy <- run_sector_pca(
  clean_list  = Canopy_clean_list,
  sector      = "Canopy",
  id_col      = "file_name",
  min_var_PC1 = 0.60,
  results_dir = "output2",
  plot_dir    = "output2"
)

```


```{r}
# ============================================================
# 2x2 BIPLOTS (Tree / Branch / Canopy) - 1200 dpi
# - 4 pannelli (QC_TR, QC_CP, FS_HF, PM_R) con lettere a–d
# - ellisse 95%
# - etichette variabili GRANDI e senza linee di collegamento
# - pannelli stessa dimensione (scale assi possono differire)
# ============================================================

library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)

# ---------------------------
# 0) Impostazioni
# ---------------------------
pops_order <- c("QC_TR","QC_CP","FS_HF","PM_R")

out_dir <- "C:/Users/rosse/OneDrive/Desktop/Lidar structure/Stat/Results/PCA_Sectors/03_biplots_1200dpi"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ---------------------------
# 1) PCA builder (stessa logica pulizia)
# ---------------------------
build_pca_for_biplot <- function(df, id_col = "file_name") {

  num <- df %>% dplyr::select(where(is.numeric))
  num[!is.finite(as.matrix(num))] <- NA

  sd_vec <- apply(num, 2, sd, na.rm = TRUE)
  num <- num[, sd_vec > 0, drop = FALSE]

  good_rows <- complete.cases(num)
  num_clean <- num[good_rows, , drop = FALSE]

  if (ncol(num_clean) < 2 || nrow(num_clean) < 3) return(NULL)

  pca <- prcomp(num_clean, center = TRUE, scale. = TRUE)
  var_expl <- (pca$sdev^2) / sum(pca$sdev^2)

  scores <- as.data.frame(pca$x[, 1:2, drop = FALSE])
  colnames(scores) <- c("PC1", "PC2")

  load <- as.data.frame(pca$rotation[, 1:2, drop = FALSE])
  colnames(load) <- c("PC1", "PC2")
  load$Variable <- rownames(load)
  rownames(load) <- NULL

  list(scores = scores, loadings = load, var = var_expl[1:2])
}

# ---------------------------
# 2) Singolo biplot (ggplot)
#    - NO linee verso le label: segment.color = NA
# ---------------------------
make_biplot_one <- function(clean_list, pop, sector_label, tag_letter) {

  df <- clean_list[[pop]]
  if (is.null(df)) return(NULL)

  pca_obj <- build_pca_for_biplot(df)
  if (is.null(pca_obj)) return(NULL)

  sc <- pca_obj$scores
  ld <- pca_obj$loadings
  v  <- pca_obj$var

  # scala frecce: proporziona ai range degli scores
  r_x <- diff(range(sc$PC1, na.rm = TRUE))
  r_y <- diff(range(sc$PC2, na.rm = TRUE))
  arrow_scale <- 0.30 * min(r_x, r_y)

  ld2 <- ld %>%
    mutate(PC1s = PC1 * arrow_scale,
           PC2s = PC2 * arrow_scale)

  # percentuali varianza
  pc1_lab <- paste0("PC1 (", round(100 * v[1], 1), "%)")
  pc2_lab <- paste0("PC2 (", round(100 * v[2], 1), "%)")

  ggplot(sc, aes(PC1, PC2)) +
    geom_point(size = 2.2, alpha = 0.75, colour = "grey45") +

    # Ellisse 95% (score-space)
    stat_ellipse(level = 0.95, linetype = "dashed", linewidth = 0.7, colour = "black") +

    # frecce loadings
    geom_segment(
      data = ld2,
      aes(x = 0, y = 0, xend = PC1s, yend = PC2s),
      inherit.aes = FALSE,
      linewidth = 0.6,
      colour = "black",
      arrow = arrow(length = unit(0.18, "cm"))
    ) +

    # label variabili GRANDI, senza segmenti (no linee)
    geom_text_repel(
      data = ld2,
      aes(x = PC1s, y = PC2s, label = Variable),
      inherit.aes = FALSE,
      size = 4.6,                 # <-- più grande
      fontface = "plain",
      colour = "black",
      segment.color = NA,         # <-- elimina le linee di collegamento
      max.overlaps = Inf,
      box.padding = 0.25,
      point.padding = 0.15
    ) +

    labs(
      title = paste0(sector_label, " sector — ", pop),
      x = pc1_lab, y = pc2_lab
    ) +

    # lettera a–d in basso a destra (dentro pannello)
    annotate("text", x = Inf, y = -Inf, label = tag_letter,
             hjust = 1.1, vjust = -0.7, size = 6) +

    scale_x_continuous(expand = expansion(mult = 0.18)) +
    scale_y_continuous(expand = expansion(mult = 0.18)) +

    coord_cartesian(clip = "off") +
    theme_classic(base_size = 15) +
    theme(
      plot.title = element_text(size = 16, face = "plain"),
      axis.title = element_text(size = 15),
      axis.text  = element_text(size = 12),
      plot.margin = margin(10, 16, 10, 10)
    )
}

# ---------------------------
# 3) Panel 2x2 + export 1200 dpi
# ---------------------------
save_biplot_panel_2x2 <- function(clean_list, sector_label, out_file,
                                  pops_order = c("QC_TR","QC_CP","FS_HF","PM_R"),
                                  dpi = 1200) {

  tags <- letters[seq_along(pops_order)]

  plist <- vector("list", length(pops_order))
  for (k in seq_along(pops_order)) {
    pop <- pops_order[k]
    plist[[k]] <- make_biplot_one(clean_list, pop, sector_label, tags[k])
  }

  # patchwork: pannelli stessa dimensione, scale assi libere
  panel <- (plist[[1]] | plist[[2]]) / (plist[[3]] | plist[[4]])

  ggsave(
    filename = out_file,
    plot = panel,
    width = 12, height = 10, units = "in",
    dpi = dpi, bg = "white"
  )
}

# ---------------------------
# 4) Esegui: TREE / BRANCH / CANOPY
# ---------------------------
save_biplot_panel_2x2(Tree_clean_list,   "Tree",
                      file.path(out_dir, "Biplot_Tree_ALL_2x2_1200dpi.png"),
                      pops_order = pops_order, dpi = 1200)

save_biplot_panel_2x2(Branch_clean_list, "Branch",
                      file.path(out_dir, "Biplot_Branch_ALL_2x2_1200dpi.png"),
                      pops_order = pops_order, dpi = 1200)

save_biplot_panel_2x2(Canopy_clean_list, "Canopy",
                      file.path(out_dir, "Biplot_Canopy_ALL_2x2_1200dpi.png"),
                      pops_order = pops_order, dpi = 1200)

```





```{r}

# -------------------------------
# REMOVE FILE EXTENSION (.txt, .mat, .ply, ecc.)
# -------------------------------

remove_ext <- function(x) gsub("\\.[^.]+$", "", x)

res_tree$scores$file_name   <- remove_ext(res_tree$scores$file_name)
res_branch$scores$file_name <- remove_ext(res_branch$scores$file_name)
res_canopy$scores$file_name <- remove_ext(res_canopy$scores$file_name)


library(dplyr)

tree_scores <- res_tree$scores %>%
  select(file_name, Pop, starts_with("Tree_PC"))

branch_scores <- res_branch$scores %>%
  select(file_name, Pop, starts_with("Branch_PC"))

canopy_scores <- res_canopy$scores %>%
  select(file_name, Pop, starts_with("Canopy_PC"))

# uniamo per file_name + Pop (albero e popolamento)
sector_scores <- tree_scores %>%
  full_join(branch_scores, by = c("file_name", "Pop")) %>%
  full_join(canopy_scores, by = c("file_name", "Pop"))

# controlliamo
dim(sector_scores)
head(sector_scores)


sector_complete <- sector_scores %>%
  filter(
    !is.na(Tree_PC1),
    !is.na(Branch_PC1),
    !is.na(Canopy_PC1)
  )

```


```{r}
make_sector_indicator <- function(res, prefix) {
  scores_df <- res$scores      # contiene file_name, Pop, Tree_PC1, Tree_PC2, ...
  var_df    <- res$variance    # contiene Pop, Component, VarExpl
  
  out_list <- list()
  
  for (pp in unique(scores_df$Pop)) {
    
    S_pop <- scores_df %>% 
      dplyr::filter(Pop == pp)
    
    # colonne di PC per questo settore (es. "Tree_PC1","Tree_PC2")
    pc_cols <- grep(paste0("^", prefix), names(S_pop), value = TRUE)
    
    if (length(pc_cols) == 0) next
    
    pcs <- as.matrix(S_pop[, pc_cols, drop = FALSE])
    m_s <- ncol(pcs)
    
    # prendo i VarExpl per questo popolamento e per quelle PC
    comp_names <- paste0("PC", seq_len(m_s))
    
    v_pop <- var_df %>% 
      dplyr::filter(Pop == pp, Component %in% comp_names) %>%
      dplyr::arrange(Component) %>%
      dplyr::pull(VarExpl)
    
    # se qualcosa va storto, salto
    if (length(v_pop) != m_s) next
    
    w <- v_pop / sum(v_pop)      # pesi normalizzati
    
    indicator <- as.numeric(pcs %*% w)
    
    # nome del punteggio di settore (TreeScore, BranchScore, CanopyScore)
    sector_name <- gsub("_PC$", "", prefix)  # "Tree_PC" -> "Tree"
    score_name  <- paste0(sector_name, "Score")
    
    S_pop[[score_name]] <- indicator
    
    out_list[[pp]] <- S_pop
  }
  
  dplyr::bind_rows(out_list)
}


```

```{r}
Tree_ind   <- make_sector_indicator(res_tree,   "Tree_PC")
Branch_ind <- make_sector_indicator(res_branch, "Branch_PC")
Canopy_ind <- make_sector_indicator(res_canopy, "Canopy_PC")

head(Tree_ind)
head(Branch_ind)
head(Canopy_ind)



```

```{r}
tree_s   <- Tree_ind   %>% select(file_name, Pop, TreeScore)
branch_s <- Branch_ind %>% select(file_name, Pop, BranchScore)
canopy_s <- Canopy_ind %>% select(file_name, Pop, CanopyScore)

sector_scores2 <- tree_s %>%
  full_join(branch_s, by = c("file_name","Pop")) %>%
  full_join(canopy_s, by = c("file_name","Pop"))

# tieni solo gli alberi con TUTTI e tre i punteggi di settore
sector_complete2 <- sector_scores2 %>%
  filter(
    !is.na(TreeScore),
    !is.na(BranchScore),
    !is.na(CanopyScore)
  )

```

```{r}
sector_z <- sector_complete2 %>%
  group_by(Pop) %>%
  mutate(
    ZTree   = as.numeric(scale(TreeScore)),
    ZBranch = as.numeric(scale(BranchScore)),
    ZCanopy = as.numeric(scale(CanopyScore))
  ) %>%
  ungroup()

head(sector_z)

```

# Integrative PCA

	The modified structural heterogeneity index (mSHI-TLS) 
The three standardised sector indicators were combined through an integrative PCA performed independently for each forest stand. This integrative PCA was conducted exclusively to the three variance-balanced and mutually orthogonal sector-level scores. For each tree i, the modified TLS-based Structural Heterogeneity Index (mSHI-TLS) was defined as the score on the first principal component of the integrative PCA.
The first integrative component was retained when explaining more than 60% of the total variance, and its orientation was constrained so that the sum of loadings was positive, ensuring that higher mSHI-TLS values consistently represented greater whole-tree structural heterogeneity.

At the stand level, the median of the tree-level mSHI-TLS distribution was used as a robust descriptor of central structural heterogeneity, while the interquartile range (IQR) was computed as a complementary measure of within-stand structural variability. Together, the median and IQR position each stand in a bivariate space describing both its central structural heterogeneity and its internal heterogeneity. The mSHI-TLS is a dimensionless, continuous index, with values centred around zero; higher values indicate greater whole-tree structural heterogeneity. 

```{r}
run_integrative_pca <- function(df, id_col = "file_name") {
  
  pops <- unique(df$Pop)
  
  scores_list <- list()
  var_list    <- list()
  load_list   <- list()
  
  for (pp in pops) {
    
    sub <- df %>% filter(Pop == pp)
    
    X <- sub %>%
      select(ZTree, ZBranch, ZCanopy) %>%
      as.matrix()
    
    if (nrow(X) < 2) next
    
    # PCA sulle 3 variabili standardizzate
    pca <- prcomp(X, center = TRUE, scale. = FALSE)
    
    var_expl <- (pca$sdev^2) / sum(pca$sdev^2)
    cum_var  <- cumsum(var_expl)
    
    # Loadings PC1
    load1 <- pca$rotation[, 1]
    
    # Orientamento (PC1 sempre positiva con complessità crescente)
    if (sum(load1, na.rm = TRUE) < 0) {
      load1      <- -load1
      pca$x[, 1] <- -pca$x[, 1]
    }
    
    # ----- SCORE INTEGRATIVO -----
    score_df <- data.frame(
      file_name      = sub[[id_col]],
      Pop            = pp,
      Integrated_PC1 = pca$x[, 1]
    )
    
    scores_list[[pp]] <- score_df
    
    # ----- VARIANZA SPIEGATA -----
    var_df <- tibble(
      Pop       = pp,
      Component = paste0("PC", seq_along(var_expl)),
      VarExpl   = var_expl,
      CumVar    = cum_var
    )
    var_list[[pp]] <- var_df
    
    # ----- LOADINGS -----
    load_df <- tibble(
      Pop         = pp,
      Variable    = c("ZTree", "ZBranch", "ZCanopy"),
      Loading_PC1 = as.numeric(load1)
    )
    load_list[[pp]] <- load_df
  }
  
  list(
    scores   = bind_rows(scores_list),
    variance = bind_rows(var_list),
    loadings = bind_rows(load_list)
  )
}
mshi_pca <- run_integrative_pca(sector_z)
```

```{r}
sector_mshi <- sector_z %>%
  left_join(mshi_pca$scores, by = c("file_name", "Pop")) %>%
  rename(mSHI_TLS = Integrated_PC1)

dim(sector_mshi)
head(sector_mshi)

write_csv(sector_mshi, file.path(output, "mSHI_TLS_trees.csv"))

```

```{r}
mshi_pop_median <- sector_mshi %>%
  group_by(Pop) %>%
  summarise(
    median_mSHI = median(mSHI_TLS, na.rm = TRUE),
    IQR_mSHI    = IQR(mSHI_TLS, na.rm = TRUE)
  )

mshi_pop_median
write_csv(mshi_pop_median, file.path(output, "mSHI_TLS_pop.csv"))
```











